

# Regression Part 1

In this workshop we will explore Simple Regression in the context of the General Linear Model (GLM). You will also have the opportunity to build some regression models where you predict an outcome variable on the basis of one predictor. You will also learn how to run model diagnostics to ensure you are not violating any key assumptions of regression.

## Overview
First off I'd like you to watch the following video which starts off by revising the basics of correlation, before examining how we build regressions models. 

&nbsp;&nbsp;

<center>

<iframe width="560" height="315" src="https://youtube.com/embed/0t-dzaEDiPg" data-external="1" frameborder="0" allowfullscreen></iframe>

</center>

&nbsp;&nbsp;

&nbsp;&nbsp;

## Simple Linear Regression
After having watched the video above, I'd like you to work through the following simple linear regression example in R. Remember to create a new `.Rproj` file to keep things organised.

### The Packages We Need

First we need to install the packages we need.  We're going to install the `tidyverse` set of packages plus a few others. The package `Hmisc` allows us to use the `rcorr()` function for calculating Pearson's r, and the `performance` package so we can test our model assumptions.  Remember, if you haven't previously installed these packages on your laptop you first need to type `install.packages("packagename")` in the console before you can call the `library()` function for that package. You *may* also need to install the package `see` to get the `performance` package working. If so, do that in the console by typing `install.packages("see")`.

```{r, message=FALSE}
library(tidyverse)
library(Hmisc)
library(performance)
```

### Import the Data

Import the dataset called `crime_dataset.csv` - this dataset contains population data, housing price index data and crime data for cities in the US.

We can use the function `head()` to display the first few rows of our dataset called "crime".

```{r, message=FALSE}
crime <- read_csv("https://raw.githubusercontent.com/george-farmer/PCHN63101/refs/heads/main/reg_pt1.csv")
head(crime)
```

### Tidy the Data

First let's do some wrangling.  There is one column that combines both City and State information. Let's separate that information out into two new columns called "City" and "State" using the function `separate()`. We'll also rename the columns to change the name of the "index_nsa" column to "House_price" and get rid of the space in the "Violent Crimes" heading.

```{r}
crime_tidied <- crime %>%
  separate(col = "City, State", into = c("City", "State")) %>%
  rename(House_price = index_nsa) %>%
  rename(Violent_Crimes = "Violent Crimes")
head(crime_tidied)
```

### Plot the Data

We might first think that as population size increases, crime rate also increases.  Let's first build a scatter plot.

```{r, warning=FALSE, message=FALSE}
crime_tidied %>%
  ggplot(aes(x = Population, y = Violent_Crimes)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() +
  theme(text = element_text(size = 13)) +
  labs(x = "Population", 
       y = "Violent Crimes")
```

### Pearson's r

This plot looks pretty interesting.  How about calculating Pearson's r?

```{r}
rcorr(crime_tidied$Population, crime_tidied$Violent_Crimes)
```

Look at the r and p-values - r is =.81 and p < .001. So ~66% of the variance in our Violent_Crimes variable is explained by our Population size variable.  Clearly there is a positive relationship between population size and the rate of violent crime. From the plot, we might conclude that the relationship is being overly influenced by crime in a small number of very large cities (top right of the plot above).  Let's exclude cities with populations greater than 2,000,000

```{r}
crime_filtered <- filter(crime_tidied, Population < 2000000)
```

Now let's redo the plot.  As there are still likely to be quite a lot of points (and thus overplotting with many points appearing roughly in the same place), we can set the alpha parameter to be < 1 in the `geom_point()` line of code. This parameter corresponds to the translucency of each point. Change it to other values to see what happens. 

```{r, warning=FALSE, message=FALSE}
crime_filtered %>%
  ggplot(aes(x = Population, y = Violent_Crimes)) + 
  geom_point(alpha = .25) + 
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() +
  theme(text = element_text(size = 13)) +
  labs(x = "Population", 
       y = "Violent Crimes")
```

And calculate Pearson's r.

```{r}
rcorr(crime_filtered$Population, crime_filtered$Violent_Crimes)
```

There is still a clear positive relationship (r=.69).  Let's build a linear model. The dataset contains a lot of data and each city appears a number of times (once each year). For our linear model, our observations need to be independent of each other so let's just focus on the year 2015. That way each city will just appear once.

First we apply our filter.

```{r}
crime_filtered <- filter(crime_filtered, Year == 2015)
```

Then we build a plot. I'm using the layer `geom_text()` to plot the City names and set the check_overlap parameter to `TRUE` to ensure the labels don't overlap.

```{r, warning=FALSE, message=FALSE}
crime_filtered %>%
  ggplot(aes(x = Population, y = Violent_Crimes, label = City)) + 
  geom_point() + 
  geom_text(nudge_y = 500, check_overlap = TRUE) + 
  geom_smooth(method = "lm", se = FALSE) + 
  xlim(0, 1800000) +
  theme_minimal() +
  theme(text = element_text(size = 13)) +
  labs(x = "Population", 
       y = "Violent Crimes")
```

This shows a clear positive linear relationship so let's work out Pearson's r.

```{r}
rcorr(crime_filtered$Population, crime_filtered$Violent_Crimes)
```

### Model the Data

Imagine we are a city planner, and we want to know by how much we think violent crimes might increase as a function of population size. In other words, we want to work out how the violent crime rate is predicted by population size.

We're going to build two linear models - one `model1` where we're using the mean of our outcome variable as the predictor, and a second `model2` where we are using Population size to predict the Violent Crimes outcome.

```{r}
model1 <- lm(Violent_Crimes ~ 1, data = crime_filtered)
model2 <- lm(Violent_Crimes ~ Population, data = crime_filtered)
```

### Checking Our Assumptions

Let's use the `check_model()` function from the performance package to check the assumptions of our model.

```{r, warning=FALSE, message=FALSE, fig.height=6}
check_model(model2, check = c("homogeneity","outliers","qq"))
```

Our dataset is small and so some of our diagnostic plots don't look great. But for now let's use the `anova()` function to see if our model with Population as the predictor is better than the one using just the mean.

```{r}
anova(model1, model2)
```

It is - the models differ and you'll see the residual sum of squares (or the error) is less in the second model (which has Population as the predictor). This means the deviation between our observed data and the regression line model `model2` is significantly less than the deviation between our observed data and the mean as a model of our data `model1`. So let's get the parameter estimates of `model2`.

### Interpreting Our Model

```{r}
summary(model2)
```

The intercept corresponds to where our regression line intercepts the y-axis, and the Population parameter corresponds to the slope of our line. We see that for every increase in population by 1 there is an extra 0.006963 increase in violent crime. 

For a city with a population of about a million, there will be about 7907 Violent Crimes. We calculate this by multiplying the estimate of our predictor (0.006963) by 1,000,000 and then adding the intercept (944.3).  This gives us 7907.3 crimes - which tallys with what you see in our regression line above. We may have a few outliers - how would you figure out what those were? Try excluding any outliers you find and re-building your model.

<center>
**End of workshop 6 materials**
</center>
